---
title: "Homework 1: Descriptive Analysis of ICE Arrests and Governor's Party"
author: "Jaye Seawright"
date: "Due: October 8, 2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(ggplot2)
```

## Homework 1: ICE Arrests and Governor's Party

**Due: Thursday, October 8, 2026 by 11:59 PM on Canvas**

There has been a great deal of discussion about the patterns of immigration arrests carried out by ICE during 2025–2026. In this assignment, we will look at data about these arrests, collected by the [Deportation Data Project](https://deportationdata.org/).

For this assignment, you are free to work individually or in groups. If you choose to work collaboratively, please include your name and the names of the people you worked with on your submitted assignment. You will not be penalized for collaborating.

---

## Data Preparation

We will use three datasets:

- `governor_data.csv`: Information about each state's governor (party, etc.)
- `ice_arrests.csv`: Individual ICE arrest records from 2025–early 2026
- `state_pops.csv`: State population estimates (2024)

We will load these and then create a state‑level dataset that includes total arrests, governor's party, and population.

### Step 1: Load and examine the data

```{r, eval=FALSE}
# Load data
governor_data <- read.csv("https://github.com/jnseawright/ps210/raw/refs/heads/main/Data/stategovs.csv")
state_pops <- read.csv("https://github.com/jnseawright/ps210/raw/refs/heads/main/Data/statepops.csv")

# ICE arrests data is provided as a zip file because the CSV is large
temp <- tempfile()
download.file("https://github.com/jnseawright/ps210/raw/refs/heads/main/Homework/ice_arrests_2025_early_2026.zip", temp)
unzip(temp, "ice_arrests_2025_early_2026.csv")
ice_arrests <- read.csv("ice_arrests_2025_early_2026.csv")
unlink(temp)  # remove the temporary file

# Look at the first few rows
head(governor_data)
head(ice_arrests)
head(state_pops)
```

### Step 2: Aggregate arrests by state

The `ice_arrests` dataset has a column `apprehension_state` indicating the state where the arrest occurred. Some records have missing state or occur outside the 50 states. We will keep only those where `apprehension_state` appears in `governor_data$State`, then count arrests per state.

```{r, eval=FALSE}
arrest_counts <- ice_arrests %>%
  filter(apprehension_state %in% governor_data$State) %>%
  group_by(apprehension_state) %>%
  summarise(arrests = n(), .groups = "drop") %>%
  rename(State = apprehension_state)

head(arrest_counts)
```

### Step 3: Add governor party and state population

Now we add governor party and population using `left_join()`. Also ensure population is numeric (it may have commas).

```{r, eval=FALSE}
# Clean population: remove commas and convert to numeric
state_pops <- state_pops %>%
  mutate(Population = parse_number(Population2024))

# Merge
state_data <- arrest_counts %>%
  left_join(governor_data %>% select(State, Party), by = "State") %>%
  left_join(state_pops %>% select(State, Population), by = "State")

# Check
head(state_data)
summary(state_data)
```

### Step 4: Filter to Democratic and Republican governors only

We exclude states with third‑party or independent governors for simplicity.

```{r, eval=FALSE}
state_data <- state_data %>%
  filter(Party %in% c("Democratic", "Republican"))

table(state_data$Party)
```

Now we have a clean dataset `state_data` with:
- `State`
- `arrests`: total ICE arrests
- `Party`: governor's party (Democratic or Republican)
- `Population`: state population (numeric)

---

## Problem 1: Descriptive Comparison (10 points)

a) **Create a bar plot** showing the total number of arrests for Democratic‑governed states and Republican‑governed states. Use `geom_bar(stat = "identity")` after summarizing by party. Add labels and a title. (3 points)

b) **Calculate the mean arrests** for Democratic‑governed states and for Republican‑governed states. Which group has a higher average? (3 points)

c) **Explain** what we learn from this descriptive comparison. What conclusions can we draw, and what further information would you want before interpreting these results causally? (4 points)

---

## Problem 2: Causal Inference and Confounding (20 points)

a) **Are there any challenges in interpreting the relationship between governor's party and ICE arrests as causal?** Work through the **four questions for causal inference** discussed in class. (8 points)

b) **Propose at least three possible confounding variables** that you would like to see explored. For each, explain why you believe it meets the criteria for a confounder (i.e., it is related to both governor's party and ICE arrests). (12 points)

---

## Problem 3: Adding Population as a Control (20 points)

We suspect that **state population** might be a confounder. Larger states have more people, potentially more arrests, and may also differ in governor party.

a) **Create a scatter plot** with `Population` on the x‑axis and `arrests` on the y‑axis, colored by `Party`. Add appropriate labels. Describe what you see. (4 points)

b) **Group states into population categories** using the `cut()` function. For example:

```{r, eval=FALSE}
state_data <- state_data %>%
  mutate(pop_group = cut(Population,
                         breaks = quantile(Population, probs = seq(0, 1, 0.2), na.rm = TRUE),
                         labels = c("Very Small", "Small", "Medium", "Large", "Very Large"),
                         include.lowest = TRUE))
```

Then create a **bar plot** showing the number of arrests by population group, with bars colored by party (`position = "dodge"`). (6 points)

c) **Interpret the plot**: Does adding information about population change the apparent relationship between party and arrests? (For example, does it *eliminate*, *reverse*, *leave unchanged*, or *complicate* the relationship?) Explain your reasoning. (5 points)

d) **Does population appear to be a confounder?** Why or why not? Use the definition of a confounder in your answer. (5 points)

---

## Problem 4: Experiments as an Alternative Approach (10 points)

a) Is this an area of research where a **social‑science experiment** makes sense as a research tool? Why or why not? (4 points)

b) If you answered **yes**, suggest an example of an experiment that would help illuminate the questions at work here. Be specific about treatment, outcome, sample, and design (lab, field, or survey). If you answered **no**, explain your reasoning for why this is a poor area for experiments. (6 points)

---

## Submission Instructions

Knit your R Markdown file to HTML or PDF, and submit it on Canvas. Make sure your code is visible and your answers are clearly written.

**Reminder:** You may work in groups, but each person must submit their own work. Include the names of any collaborators at the top of your assignment.

**Due: Thursday, October 8, 2026, 11:59 PM.**