---
title: "9: Correlation and Bivariate Regression"
subtitle: "Empirical Methods"
author: "<large>Jaye Seawright</large>"
institute: "<small>Northwestern Political Science</small>" 
date: "Oct. 14, 2025"
output: 
  xaringan::moon_reader:
    css: xaringan-themer.css
  
---
class: center, middle

```{css, echo=FALSE}
pre {
  max-height: 400px;
  overflow-y: auto;
}

pre[class] {
  max-height: 200px;
}
```

```{r, load_refs, include=FALSE, cache=FALSE}
# Initializes
library(RefManageR)

library(ggplot2)
library(dplyr)
library(readr)
library(nlme)
library(jtools)
library(mice)
library(knitr)
library(DiagrammeR)
library(extrafont)
loadfonts(device="win")

BibOptions(check.entries = FALSE,
           bib.style = "authoryear", # Bibliography style
           max.names = 3, # Max author names displayed in bibliography
           sorting = "nyt", #Name, year, title sorting
           cite.style = "authoryear", # citation style
           style = "markdown",
           hyperlink = FALSE,
           dashed = FALSE)

```
```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer,MnSymbol)
style_mono_accent(
  base_color = "#1c5253",
  header_font_google = google_font("Josefin Sans"),
  text_font_google   = google_font("Montserrat", "300", "300i"),
  code_font_google   = google_font("Fira Mono"),
  text_font_size = "1.6rem"
)

knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```


### Recall where we left off

We learned how to test whether a difference between two groups is statistically significant.

But what if our independent variable is **quantitative**, not a simple group membership?

And what if we need to account for **multiple variables** at once?

---

### Today: Bivariate Regression

- The simplest case: one quantitative independent variable, one quantitative dependent variable.
- We'll learn how to fit a line to data and interpret its slope.

---

### Regression Analysis

-   Regression analysis is a tool that fits a line to a scatter plot.

---

Remember from your studies that the algebraic formula for a line is: $Y = mX + b$.

---

### Regression Analysis

When we pivot from geometry to regression, two important changes:

1.  Instead of $m$, we usually call the slope $\beta$ or $b$, and instead of $b$, we usually call the intercept $\alpha$, $a$, or sometimes $\beta_0$.

2.  We add a term at the end of the equation for error/noise above and below the line. This term is called $e_i$ or $\epsilon_i$.

---

-   How do we get numbers for $\alpha$ and $\beta$ in regression analysis?

-   Data plus calculus! The line is chosen to minimize the sum of squared vertical distances from the points to the line (ordinary least squares).

---

### Example: Attitudes and Policy Preferences

We'll use data from the 2024 American National Election Studies (ANES).

Let's look at two variables:

- **transbathroom**: Support for laws requiring transgender individuals to use bathrooms matching their sex assigned at birth (1 = favor a great deal, 5 = oppose a great deal; lower = more conservative on this issue).
- **carbonregulations**: Support for government regulation of carbon emissions (1 = favor a great deal, 5 = oppose a great deal; lower = more supportive of regulation).

```{r, echo = TRUE, out.width="100%", fig.retina = 1, fig.align='center'}
anes2024 <- read.csv("https://raw.githubusercontent.com/jnseawright/ps210/refs/heads/main/Data/anes2024.csv")

anes2024$transbathroom <- ifelse(anes2024$V241372x < 0, NA, anes2024$V241372x)

table(anes2024$transbathroom)
```

---

```{r, echo = TRUE, out.width="100%", fig.retina = 1, fig.align='center'}
anes2024$carbonregulations <- ifelse(anes2024$V242324x < 0, NA, anes2024$V242324x)

table(anes2024$carbonregulations)
```

---

### Visualizing the Relationship

```{r, echo = TRUE, out.width="100%", fig.retina = 1, fig.align='center'}
library(ggplot2)

transcarbonplot <- anes2024 %>% 
  ggplot(aes(x = transbathroom, y = carbonregulations)) + 
  geom_point(color = "#69b3a2") +
  theme_minimal()
```

---

```{r, echo = TRUE, out.width="55%", fig.retina = 1, fig.align='center'}
transcarbonplot
```

The points overlap a lot because both variables are discrete. Let's jitter them.

---

```{r, echo = TRUE, out.width="100%", fig.retina = 1, fig.align='center'}
transcarbonjitterplot <- anes2024 %>% 
  ggplot(aes(x = transbathroom, y = carbonregulations)) + 
  geom_point(color = "#69b3a2") +
  geom_jitter() +
  theme_minimal()
```

---

```{r, echo = TRUE, out.width="55%", fig.retina = 1, fig.align='center'}
transcarbonjitterplot
```

---
### Correlation: Quantifying the Relationship

Before we fit a line, let's measure the strength and direction of the relationship.

**Correlation coefficient** ($r$):
- Ranges from -1 to +1
- Positive $r$ = as X increases, Y tends to increase
- Negative $r$ = as X increases, Y tends to decrease
- $r = 0$ = no linear relationship
- Closer to ±1 = stronger relationship

---

```{r, echo = TRUE, out.width="100%", fig.retina = 1, fig.align='center'}
# Calculate the correlation between transbathroom and carbonregulations
cor(anes2024$transbathroom, anes2024$carbonregulations, use = "complete.obs")
```

The correlation is about **0.45**, which is a pretty strong positive relationship for individual-level survey data.

---

### Visualizing Correlation

```{r, echo = TRUE, out.width="100%", fig.retina = 1, fig.align='center'}
# Install if needed: install.packages("GGally")
library(GGally)

# Create a scatterplot with correlation displayed
corplot <- anes2024 %>% 
  select(transbathroom, carbonregulations) %>%
  ggpairs(
   lower = list(
      continuous = wrap("points", position = position_jitter(width = 0.2, height = 0.2), alpha = 0.5)
   )
  ) +
  theme_minimal()
```

---

```{r, echo = FALSE, out.width="70%", fig.retina = 1, fig.align='center'}
corplot
```

---

### Correlation vs. Regression

**Correlation** tells us the strength and direction of a linear relationship.

**Regression** will give us a line that predicts Y from X, with a slope we can interpret substantively.

Correlation provides a quick summary, and regression provides an often more useful descriptive, analytic, and forecasting tool.

* The slope in a bivariate regression is related to the correlation: $\beta = r * \frac{s_Y}{s_X}$.

---

### The Bivariate Regression Model

$\scriptsize CarbonRegulations_{i} = \alpha + \beta TransBathroom_{i} + \epsilon_{i}$

We estimate $\alpha$ and $\beta$ using `lm()` in R.

---

```{r, echo = TRUE, out.width="100%", fig.retina = 1, fig.align='center'}
transcarbonlm <- lm(carbonregulations ~ transbathroom, data = anes2024)

summary(transcarbonlm)
```

---

### Interpreting the Output

- The intercept ($\alpha$) ≈ 1.02: the predicted carbon regulations support when transbathroom = 0 (a value one step below the bottom of the scale, but necessary to structure the geometry).
- The slope ($\beta$) ≈ 0.37: for each one-unit increase in support for bathroom laws (i.e., moving one step toward opposition), predicted support for carbon regulations *increases* by 0.37 units.

---

### Interpreting the Output

- The relationship is positive: people who are more conservative on transgender bathroom laws also tend to be less supportive of carbon regulation (higher values = more opposition).
- The p-value for transbathroom is very small (<2e-16), so the relationship is statistically significant.
- This means that if the true slope were zero, we would virtually never see a slope this large just by random chance.

---

### Visualizing the Regression Line

```{r, echo = TRUE, out.width="100%", fig.retina = 1, fig.align='center'}
transcarbonlmplot <- anes2024 %>% 
  ggplot(aes(x = transbathroom, y = carbonregulations)) + 
  geom_point(color = "#69b3a2") +
  geom_jitter() +
  geom_smooth(method = lm, color = "red", se = TRUE) +  
  theme_minimal()
```

---

```{r, echo = TRUE, out.width="55%", fig.retina = 1, fig.align='center', warning=FALSE, message=FALSE}
transcarbonlmplot
```

The red line is the best-fit line; the shaded band is the 95% confidence interval for the line.

---

### Summary: Bivariate Regression

- We can now quantify the relationship between two quantitative variables.
- The slope tells us the direction and strength of the association.
- We can test whether the slope is significantly different from zero.

**But**—this is just an association. It doesn't account for other variables that might be confounders.

Next time: **Multivariate Regression**—adding control variables to get closer to causal estimates.

---

### What's Next?

We saw that attitudes on transgender bathroom laws correlate with attitudes on carbon regulation.

But is that relationship **causal**, or does it reflect something else—like partisanship?

In the next session, we'll add a control variable (party identification) and see what happens.

We'll also learn when adding a control is **good**, **irrelevant**, or actually **harmful**.