---
title: "2: Data"
subtitle: "Empirical Methods"
author: "<large>Jaye Seawright</large>"
institute: "<small>Northwestern Political Science</small>" 
date: "Sept. 29, 2026"
output: 
  xaringan::moon_reader:
    css: xaringan-themer.css
  
---
class: center, middle

```{css, echo=FALSE}
pre {
  max-height: 400px;
  overflow-y: auto;
}

pre[class] {
  max-height: 200px;
}
```

```{r, load_refs, include=FALSE, cache=FALSE}
# Initializes
library(RefManageR)

library(ggplot2)
library(dplyr)
library(readr)
library(nlme)
library(jtools)
library(mice)
library(knitr)

BibOptions(check.entries = FALSE,
           bib.style = "authoryear", # Bibliography style
           max.names = 3, # Max author names displayed in bibliography
           sorting = "nyt", #Name, year, title sorting
           cite.style = "authoryear", # citation style
           style = "markdown",
           hyperlink = FALSE,
           dashed = FALSE)

```
```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer,MnSymbol)
style_mono_accent(
  base_color = "#1c5253",
  header_font_google = google_font("Josefin Sans"),
  text_font_google   = google_font("Montserrat", "300", "300i"),
  code_font_google   = google_font("Fira Mono"),
  text_font_size = "1.6rem"
)

usterrorism <- read.csv("https://raw.githubusercontent.com/jnseawright/ps210/refs/heads/main/Data/usterrorism.csv")

```
### Today's Goals

By the end of this session, you should be able to:

1. Identify the key strengths and limitations of a real-world data source
2. Explain how reporting practices, conceptual definitions, and invisibility bias shape what data capture
3. Use R to explore a dataset with `head()`, `table()`, `filter()`, and `ggplot()`
4. Interpret a distribution of event counts across time
5. Explain why data creation is often the most consequential step in a research project

---
### A perennially newsworthy question: how much political violence is there in the U.S., and what are its sources?

---

Let's look at the [Global Terrorism Database](https://www.start.umd.edu/data-tools/GTD).

---

What are the strengths and weaknesses of this data source?

---
### Issues to Consider

* Reporting
* Definitions of concepts (and who has the power to impose them)
* What kinds of relevant phenomena would be invisible?

---

In order to look at the data, I'm going to use a statistical analysis environment called R.

---

You can use R for free without installing anything at [Posit Cloud](https://posit.cloud/).

---
```{r, echo = TRUE, out.width="100%", fig.retina = 1, fig.align='center'}

head(usterrorism)

```

---
```{r, echo = TRUE, out.width="100%", fig.retina = 1, fig.align='center'}

table(usterrorism$nkill)

```

---
```{r, echo = TRUE, out.width="100%", fig.retina = 1, fig.align='center'}

usterrorism %>% filter(nkill==1385)

```

---
```{r, echo = TRUE, out.width="100%", fig.retina = 1, fig.align='center', fig.alt="R code and resulting boxplot: distribution of deaths per U.S. terrorism incident by year on a log scale. Most years show medians near zero with high-fatality outliers; 2001 is a dramatic outlier due to the 9/11 attacks."}

ggplot(usterrorism, aes(x = factor(iyear), y = nkill)) +
    geom_boxplot(fill = "grey", outlier.color = "black", outlier.shape = 1) +
    scale_y_continuous(trans = "log1p") +  # This transforms the y-axis to log1p scale
    labs(
        title = "Distribution of Deaths per Terrorism Incident by Year (Log Scale)",
        x = "Year",
        y = "Number of Deaths (Log Scale)"
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

---
```{r, echo = FALSE, out.width="70%", fig.retina = 1, fig.align='center', fig.alt="Boxplot: distribution of deaths per U.S. terrorism incident by year on a log scale. Most years show medians near zero with high-fatality outliers; 2001 is a dramatic outlier due to the 9/11 attacks."}

ggplot(usterrorism, aes(x = factor(iyear), y = nkill)) +
    geom_boxplot(fill = "grey", outlier.color = "black", outlier.shape = 1) +
    scale_y_continuous(trans = "log1p") +  # This transforms the y-axis to log1p scale
    labs(
        title = "Distribution of Deaths per Terrorism Incident by Year (Log Scale)",
        x = "Year",
        y = "Number of Deaths (Log Scale)"
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

---
```{r, echo = TRUE, out.width="100%", fig.retina = 1, fig.align='center'}

table(usterrorism$gname)

```

---

What is the [Phineas Priesthood](https://www.splcenter.org/resources/extremist-files/phineas-priesthood/)?

---

What about [White Rabbit Three Percent Illinois Patriot Freedom Fighters Militia](https://www.cdse.edu/Portals/124/Documents/casestudies/case-study-emily-hari.pdf)?

---
### Conclusions

1. Consider the perspective and limitations of data sources.

2. Quantitative data are often easier to work with than qualitative data, but may be less rewarding.

3. With enough work, it's usually possible to find or create data to learn an answer to your question.

4. Data creation can be where most of the work on a project takes place, and everything else usually depends on this step.