---
title: "2: Data"
subtitle: "Empirical Methods"
author: "<large>J. Seawright</large>"
institute: "<small>Northwestern Political Science</small>" 
date: "Sept. 18, 2025"
output: 
  xaringan::moon_reader:
    css: xaringan-themer.css
  
---
class: center, middle

```{css, echo=FALSE}
pre {
  max-height: 400px;
  overflow-y: auto;
}

pre[class] {
  max-height: 200px;
}
```

```{r, load_refs, include=FALSE, cache=FALSE}
# Initializes
library(RefManageR)

library(ggplot2)
library(dplyr)
library(readr)
library(nlme)
library(jtools)
library(hrbrthemes)
library(mice)
library(knitr)

BibOptions(check.entries = FALSE,
           bib.style = "authoryear", # Bibliography style
           max.names = 3, # Max author names displayed in bibliography
           sorting = "nyt", #Name, year, title sorting
           cite.style = "authoryear", # citation style
           style = "markdown",
           hyperlink = FALSE,
           dashed = FALSE)

```
```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer,MnSymbol)
style_mono_accent(
  base_color = "#1c5253",
  header_font_google = google_font("Josefin Sans"),
  text_font_google   = google_font("Montserrat", "300", "300i"),
  code_font_google   = google_font("Fira Mono"),
  text_font_size = "1.6rem"
)

usterrorism <- read.csv("https://raw.githubusercontent.com/jnseawright/ps210/refs/heads/main/Data/usterrorism.csv")

```
### A perennially newsworthy question: how much political violence is there in the U.S., and what are its sources?

---

Let's look at the [Global Terrorism Database](https://www.start.umd.edu/data-tools/GTD).

---

In order to look at the data, I'm going to use a statistical analysis environment called R.

---

You can use R for free without installing anything at [Posit Cloud](https://posit.cloud/).

---
```{r, echo = TRUE, out.width="100%", fig.retina = 1, fig.align='center'}

head(usterrorism)

```

---
```{r, echo = TRUE, out.width="100%", fig.retina = 1, fig.align='center'}

table(usterrorism$nkill)

```

---
```{r, echo = TRUE, out.width="100%", fig.retina = 1, fig.align='center'}

usterrorism %>% filter(nkill==1385)

```

---
```{r, echo = TRUE, out.width="100%", fig.retina = 1, fig.align='center'}

ggplot(usterrorism, aes(x = factor(iyear), y = nkill)) +
    geom_boxplot(fill = "grey", outlier.color = "black", outlier.shape = 1) +
    scale_y_continuous(trans = "log1p") +  # This transforms the y-axis to log1p scale
    labs(
        title = "Distribution of Deaths per Terrorism Incident by Year (Log Scale)",
        x = "Year",
        y = "Number of Deaths (Log Scale)"
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

---
```{r, echo = FALSE, out.width="70%", fig.retina = 1, fig.align='center'}

ggplot(usterrorism, aes(x = factor(iyear), y = nkill)) +
    geom_boxplot(fill = "grey", outlier.color = "black", outlier.shape = 1) +
    scale_y_continuous(trans = "log1p") +  # This transforms the y-axis to log1p scale
    labs(
        title = "Distribution of Deaths per Terrorism Incident by Year (Log Scale)",
        x = "Year",
        y = "Number of Deaths (Log Scale)"
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

---
```{r, echo = TRUE, out.width="100%", fig.retina = 1, fig.align='center'}

table(usterrorism$gname)

```

---

What is the [Phineas Priesthood](https://www.splcenter.org/resources/extremist-files/phineas-priesthood/)?

---

What about [White Rabbit Three Percent Illinois Patriot Freedom Fighters Militia](https://www.cdse.edu/Portals/124/Documents/casestudies/case-study-emily-hari.pdf)?

---
### Conclusions

1. Consider the perspective and limitations of data sources.

2. Quantitative data are often easier to work with than qualitative data, but may be less rewarding.

3. With enough work, it's usually possible to find or create data to learn an answer to your question.