---
title: "Lab 2: Your Turn Practice Exercise"
author: "Jaye Seawright"
date: "Oct. 30, 2025"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(ggplot2)
library(modelsummary)
```

## Instructions

This is your chance to practice regression skills with the **same datasets you used in Lab 1**. Work through each step below. Ask a neighbor or TA if you get stuck.

**You don't need to submit this for a grade.** It's just practice, but keeping your code will help with Homework 2.

---

## The Research Question

Does a county's support for Donald Trump in the 2024 election affect the amount of federal health‑care grant funding it receives in 2025?

We have the same datasets from Lab 1:

- `hrsa_grants.csv`: Federal health‑care grants awarded to counties since 2018
- `countypres.csv`: County‑level presidential election results (2024)

---

## Step 1: Load and Prepare the Data

The code below repeats the steps from Lab 1. Run it to create the analysis dataset.

```{r, eval=FALSE}
# Load data
grants <- read.csv("https://raw.githubusercontent.com/jnseawright/ps210/main/Data/hrsa_grants.csv")
counties <- read.csv("https://raw.githubusercontent.com/jnseawright/ps210/main/Data/countypres.csv")

# Keep only 2025 grants
grants_2025 <- grants %>% filter(Fiscal.Year == 2025)

# Keep only Trump 2024 votes (total mode) and calculate Trump vote share
counties_2024 <- counties %>%
  filter(year == 2024 & party == "REPUBLICAN" & mode == "TOTAL") %>%
  mutate(trump_share = candidatevotes / totalvotes)

# Clean county names in grants dataset
grants_2025 <- grants_2025 %>%
  mutate(
    county_clean = str_remove_all(County,
                                   regex(" County| City| Borough| Parish| Census Area",
                                         ignore_case = TRUE)),
    county_clean = str_trim(county_clean),
    county_clean = str_to_upper(county_clean)
  )

# Merge grants with election data
grants_merged <- left_join(
  grants_2025,
  counties_2024,
  by = c("State" = "state_po", "county_clean" = "county_name")
)

# Keep only relevant variables and remove rows with missing Trump share
grants_analysis <- grants_merged %>%
  select(Fiscal.Year, State, County, Funding, trump_share, totalvotes) %>%
  filter(!is.na(trump_share))
```

**Your Task:** After running the code, examine the resulting dataset with `head()` and `summary()`. Write down:

1. How many observations (grants) are in `grants_analysis`?
2. What is the range of `Funding`? Of `trump_share`?
3. Why might we want to use `log(Funding)` in a regression? (Look at the distribution.)

```{r, eval=FALSE}
# Your code here

# Your observations (as comments)
```

---

## Step 2: Visualize the Relationship

**Your Task:** Create a scatter plot of `Funding` (y) against `trump_share` (x). Add a linear smooth (`geom_smooth(method = "lm")`). Add appropriate labels.

```{r, eval=FALSE}
# Write your code here

```

**Interpret:** What does the plot suggest about the relationship? Does it appear positive, negative, or flat? Are there any outliers? Does either variable look like it could use a log transformation?

---

## Step 3: Bivariate Regression

**Your Task:** Fit a linear model predicting `Funding` from `trump_share`, using any log transformations, etc., that you concluded might be necessary based on the plot. Display the results with `summary()` and with `modelsummary()`.

```{r, eval=FALSE}
# Fit model
model_bivar <- lm(________ ~ ________, data = grants_analysis)

# Summary
summary(model_bivar)

# modelsummary
modelsummary(model_bivar, statistic = "p.value")
```

**Interpret:** 
- What is the slope? Is it positive or negative?
- Is it statistically significant? (Look at the p‑value.)
- Substantively, what does the slope mean? (Hint: For a 1‑percentage‑point increase in Trump share, how much does funding change in dollars?)

---

## Step 4: Add a Control Variable

You suspect that **county population** might be a confounder: larger counties tend to vote differently and also receive more grants (simply because they have more people and organizations). The `totalvotes` variable is a reasonable proxy for population.

**Your Task:** Fit a multivariate model with both `trump_share` and `totalvotes`. Use `modelsummary()` to compare with the bivariate model.

```{r, eval=FALSE}
# Multivariate model
model_control <- lm(________ ~ ________ + ________, data = grants_analysis)

# Compare models
modelsummary(list("Bivariate" = model_bivar, "With Population" = model_control),
             statistic = "p.value")
```

**Interpret:**
- How did the coefficient for `trump_share` change after adding `totalvotes`? Did it get closer to zero? Farther?
- Is `totalvotes` a statistically significant predictor?
- Does `totalvotes` appear to be a confounder? Why or why not? (Recall the definition: a confounder is related to both the treatment and the outcome, and including it changes the treatment coefficient.)

---

## Step 5: Reflect

Answer in a few sentences:

1. What was the most challenging part of this exercise?
2. What did you learn about the relationship between Trump support and grant funding?
3. How might these skills apply to Homework 2? (Homework 2 will ask you to do exactly this type of analysis, interpret the results, and discuss causal inference.)

```{r, eval=FALSE}
# Write your reflections here
```

---

## Done!

You've practiced:
- Loading and preparing the grants/election data
- Visualizing the relationship
- Fitting bivariate and multivariate regressions
- Interpreting coefficients and confounding

These are exactly the skills you'll need for Homework 2. Keep this file as a reference!