---
title: "Machine Learning"
subtitle: "Empirical Methods"
author: "<large>Jaye Seawright</large>"
institute: "<small>Northwestern Political Science</small>" 
date: "Nov. 10, 2026"
output: 
  xaringan::moon_reader:
    css: xaringan-themer.css
  
---
class: center, middle

```{css, echo=FALSE}
pre {
  max-height: 400px;
  overflow-y: auto;
}

pre[class] {
  max-height: 200px;
}
```

```{r, load_refs, include=FALSE, cache=FALSE}
# Initializes the bibliography
library(RefManageR)

library(ggplot2)
library(dplyr)
library(readr)
library(nlme)
library(jtools)
library(mice)
library(knitr)

BibOptions(check.entries = FALSE,
           bib.style = "authoryear", # Bibliography style
           max.names = 3, # Max author names displayed in bibliography
           sorting = "nyt", #Name, year, title sorting
           cite.style = "authoryear", # citation style
           style = "markdown",
           hyperlink = FALSE,
           dashed = FALSE)
#myBib <- ReadBib("assets/myBib.bib", check = FALSE)
# Note: don't forget to clear the knitr cache to account for changes in the
# bibliography.

knitr::opts_chunk$set(warning = FALSE, message = FALSE) 

qogsmall <- read.csv("https://github.com/jnseawright/ps210/raw/refs/heads/main/Data/qogsmall.csv")
```
```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer,MnSymbol)
style_mono_accent(
  base_color = "#1c5253",
  header_font_google = google_font("Josefin Sans"),
  text_font_google   = google_font("Montserrat", "300", "300i"),
  code_font_google   = google_font("Fira Mono"),
  text_font_size = "1.6rem"
)
```

### Today's Goals

By the end of this session, you should be able to:

1. Explain why assuming linearity limits what we can learn from data, and describe how LOESS addresses this
2. Describe how CART builds a decision tree through recursive binary splits
3. Interpret a CART tree: identify the root split, trace paths to terminal nodes, and read predicted values
4. Explain how random forests improve on a single CART tree through bagging and variable randomization
5. Interpret a minimum depth distribution plot to identify the most important predictors in a random forest

---
### A (Quick) Historical Journey

A lot of social science works with statistical models that assume the whole shape of the equation they work with, like regression did.

---
```{r, echo = TRUE, out.width="90%", fig.retina = 1, fig.align='center'}
democplot.1 <- ggplot(qogsmall, aes(x=year, y=democracy)) +
  geom_point() +
  geom_smooth(method=lm , color="red", fill="#69b3a2", se=TRUE) +
  theme_minimal()
```

---
```{r, echo = TRUE, out.width="90%", fig.retina = 1, fig.align='center', fig.alt="Scatterplot of year (x-axis) vs democracy score (y-axis) from the Quality of Government dataset. Individual country-years are plotted as points. A red linear regression line with a green 95% confidence band shows a slight upward trend in democracy over time, but considerable scatter around the line."}

democplot.1
```

---

But it turns out that we can learn a lot more from the data, if we're flexible.

---
```{r, echo = TRUE, out.width="90%", fig.retina = 1, fig.align='center'}
democplot.2 <- ggplot(qogsmall, aes(x=year, y=democracy)) +
  geom_point() +
  geom_smooth(method=loess , color="red", fill="#69b3a2", se=TRUE) +
  theme_minimal()
```
---
```{r, echo = TRUE, out.width="90%", fig.retina = 1, fig.align='center', fig.alt="Same scatterplot of year vs democracy score, but now with a LOESS (locally estimated scatterplot smoothing) curve in red and a green 95% confidence band. Unlike the linear model, the flexible curve can reveal non-linear patterns — showing how democracy levels rise and fall at different rates across different time periods."}
democplot.2
```

---

What we're doing in this plot is a LOESS (locally estimated scatterplot smoothing) regression, which fits different regressions to different parts of the relationship and then averages those together to generate a smooth, potentially curved model based on learning the shape from the data.

---

LOESS is great for one or two predictors. But what if we have many variables that might interact in complex ways? That's where tree-based methods shine.

---
### CART

Optimally predicting $Y$ based on $\mathbb{X}$, without assumptions of
additivity, linearity, etc.

---
### CART

1.  Start with the set of all cases, i.e., the root node.

2.  Search all values of each variable in $\mathbb{X}$ for the binary
    split that maximizes homogeneity of $Y$ for cases on each side of
    the split.

3.  If homogeneity of $Y$ is sufficient or if the remaining sets of
    cases as the new final nodes are too small, stop. Otherwise, repeat
    step 2 for each of the current last-generation nodes.

---
### CART: Homogeneity

Homogeneity in this context means that cases in a node are as similar as possible on the outcome variable (e.g., democracy). The algorithm searches for splits that make the two resulting groups more internally consistent.

---
### An Example

```{r, echo = FALSE, out.width="90%", fig.retina = 1, fig.align='center', fig.alt="A CART decision tree with five terminal nodes. Root split: radionews < 4.5. Right branch (high radio news exposure) predicts outcome = 1. Left branch splits on intefficacy < 2.5 — higher internal political efficacy (≥ 2.5) predicts 0.59. Lower efficacy branch splits on simpletreat < 0.5 (untreated cases split by edlevel < 7.5: terminal nodes 0.29 and 0.58; treated cases split by edlevel ≥ 7.5: terminal nodes 0.28 and 0.76). Higher education among treated respondents produces the highest predicted outcome."}
library(knitr)
include_graphics("images/experimentcart.pdf")
```

---
```{r, echo = TRUE, out.width="90%", fig.retina = 1, fig.align='center'}
library(rpart)
```

---
```{r, echo = TRUE, out.width="90%", fig.retina = 1, fig.align='center'}
dem.cart <- rpart(democracy ~ corruption+genderequity+percapitagdp+refugees+centralbankindependence+imports+ecologicalfootprint+globalization, data=qogsmall, na.action=na.omit)
```

---
```{r, echo = TRUE, out.width="90%", fig.retina = 1, fig.align='center', fig.alt="CART decision tree predicting democracy scores from corruption, gender equity, per capita GDP, refugees, central bank independence, imports, ecological footprint, and globalization. The tree shows recursive binary splits; variables closer to the root are stronger predictors. Terminal nodes display the predicted democracy value for cases that follow that path."}
library(rpart.plot)
rpart.plot(dem.cart)
```

---
### Random Forests

1.  Sample randomly from the entire data.
2.  Run CART, selecting a random subset of data points and variables at each decision node.
3.  Repeat several times, and find a way to average the results
    together.

---
```{r, echo = TRUE, out.width="90%", fig.retina = 1, fig.align='center'}
library(randomForest)
```

---
```{r, echo = TRUE, out.width="90%", fig.retina = 1, fig.align='center'}
dem.rf <- randomForest(democracy ~ corruption+genderequity+percapitagdp+refugees+centralbankindependence+imports+ecologicalfootprint+globalization, data=qogsmall, na.action=na.omit)

dem.rf
```

---
```{r, echo = TRUE, out.width="90%", fig.retina = 1, fig.align='center'}
library(randomForestExplainer)
```

---
```{r, echo = TRUE, out.width="90%", fig.retina = 1, fig.align='center', fig.alt="Minimum depth distribution plot from the random forest. Each bar represents a predictor variable; bar length reflects mean minimal depth across trees (shorter = more important, appears closer to the root more often). Corruption and per capita GDP have the smallest mean minimal depths, identifying them as the strongest predictors of democracy in this model."}
dem.mindepth <- min_depth_distribution(dem.rf)
plot_min_depth_distribution(dem.mindepth)
```


---
```{r, echo = FALSE, out.width="70%", fig.retina = 1, fig.align='center', fig.alt="Minimum depth distribution plot from the random forest (display version). Corruption and per capita GDP appear with the smallest mean minimal depths, indicating they are the most important predictors of democracy; other variables such as globalization and gender equity appear deeper in trees on average."}
plot_min_depth_distribution(dem.mindepth)
```

---
### What This Plot Means

* Variables that appear closer to the root (smaller minimal depth) are more important.

* This plot shows that corruption and per capita GDP are the strongest predictors of democracy in this model.

---

```{r, echo = TRUE, out.width="90%", fig.retina = 1, fig.align='center', fig.alt="Minimum depth interaction plot from the random forest. Each cell shows how frequently a pair of predictor variables co-occurs in the same branches across trees, measured by their mean minimal depth of interaction. Pairs with lower mean interaction depth (darker cells) interact more strongly. Corruption and per capita GDP are likely among the most prominent interacting pairs."}
plot_min_depth_interactions(dem.rf)
```

---

```{r, echo = FALSE, out.width="70%", fig.retina = 1, fig.align='center', fig.alt="Minimum depth interaction plot (display version). Shows pairwise variable interactions in the random forest; pairs appearing together near the root of trees more frequently have lower interaction depth scores. This identifies which combinations of predictors jointly drive democracy scores in the model."}
plot_min_depth_interactions(dem.rf)
```

---
### What This Plot Means

This shows which pairs of variables frequently work together in the trees. For example, corruption and GDP might jointly predict democracy.

---
### What Is Machine Learning For?

* Especially prediction and forecasting: which kinds of countries are most likely to experience a coup next year?
* Can be a powerful tool for complex description problems: what topics dominate congressional speeches during conflicts?
* Researchers use machine learning to solve many kinds of challenges in causal inference: LASSO (a variant of regression) for variable selection, forests for estimating heterogeneous treatment effects
