---
title: "Machine Learning"
subtitle: "Empirical Methods"
author: "<large>Jaye Seawright</large>"
institute: "<small>Northwestern Political Science</small>" 
date: "Oct. 28, 2025"
output: 
  xaringan::moon_reader:
    css: xaringan-themer.css
  
---
class: center, middle

```{css, echo=FALSE}
pre {
  max-height: 400px;
  overflow-y: auto;
}

pre[class] {
  max-height: 200px;
}
```

```{r, load_refs, include=FALSE, cache=FALSE}
# Initializes the bibliography
library(RefManageR)

library(ggplot2)
library(dplyr)
library(readr)
library(nlme)
library(jtools)
library(mice)
library(knitr)

BibOptions(check.entries = FALSE,
           bib.style = "authoryear", # Bibliography style
           max.names = 3, # Max author names displayed in bibliography
           sorting = "nyt", #Name, year, title sorting
           cite.style = "authoryear", # citation style
           style = "markdown",
           hyperlink = FALSE,
           dashed = FALSE)
#myBib <- ReadBib("assets/myBib.bib", check = FALSE)
# Note: don't forget to clear the knitr cache to account for changes in the
# bibliography.

knitr::opts_chunk$set(warning = FALSE, message = FALSE) 

qogsmall <- read.csv("https://github.com/jnseawright/ps210/raw/refs/heads/main/Data/qogsmall.csv")
```
```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer,MnSymbol)
style_mono_accent(
  base_color = "#1c5253",
  header_font_google = google_font("Josefin Sans"),
  text_font_google   = google_font("Montserrat", "300", "300i"),
  code_font_google   = google_font("Fira Mono"),
  text_font_size = "1.6rem"
)
```

### A (Quick) Historical Journey

A lot of social science works with statistical models that assume the whole shape of the equation they work with, like regression did.

---
```{r, echo = TRUE, out.width="90%", fig.retina = 1, fig.align='center'}
democplot.1 <- ggplot(qogsmall, aes(x=year, y=democracy)) +
  geom_point() +
  geom_smooth(method=lm , color="red", fill="#69b3a2", se=TRUE) +
  theme_minimal()
```

---
```{r, echo = TRUE, out.width="90%", fig.retina = 1, fig.align='center'}

democplot.1
```

---

But it turns out that we can learn a lot more from the data, if we're flexible.

---
```{r, echo = TRUE, out.width="90%", fig.retina = 1, fig.align='center'}
democplot.2 <- ggplot(qogsmall, aes(x=year, y=democracy)) +
  geom_point() +
  geom_smooth(method=loess , color="red", fill="#69b3a2", se=TRUE) +
  theme_minimal()
```
---
```{r, echo = TRUE, out.width="90%", fig.retina = 1, fig.align='center'}
democplot.2
```

---

What we're doing in this plot is a LOESS (locally estimated scatterplot smoothing) regression, which fits different regressions to different parts of the relationship and then averages those together to generate a smooth, potentially curved model based on learning the shape from the data.

---

LOESS is great for one or two predictors. But what if we have many variables that might interact in complex ways? That's where tree-based methods shine.

---
### CART

Optimally predicting $Y$ based on $\mathbb{X}$, without assumptions of
additivity, linearity, etc.

---
### CART

1.  Start with the set of all cases, i.e., the root node.

2.  Search all values of each variable in $\mathbb{X}$ for the binary
    split that maximizes homogeneity of $Y$ for cases on each side of
    the split.

3.  If homogeneity of $Y$ is sufficient or if the remaining sets of
    cases as the new final nodes are too small, stop. Otherwise, repeat
    step 2 for each of the current last-generation nodes.

---
### CART: Homogeneity

Homogeneity in this context means that cases in a node are as similar as possible on the outcome variable (e.g., democracy). The algorithm searches for splits that make the two resulting groups more internally consistent.

---
### An Example

```{r, echo = FALSE, out.width="90%", fig.retina = 1, fig.align='center'}
library(knitr)
include_graphics("images/experimentcart.pdf")
```

---
```{r, echo = TRUE, out.width="90%", fig.retina = 1, fig.align='center'}
library(rpart)
```

---
```{r, echo = TRUE, out.width="90%", fig.retina = 1, fig.align='center'}
dem.cart <- rpart(democracy ~ corruption+genderequity+percapitagdp+refugees+centralbankindependence+imports+ecologicalfootprint+globalization, data=qogsmall, na.action=na.omit)
```

---
```{r, echo = TRUE, out.width="90%", fig.retina = 1, fig.align='center'}
library(rpart.plot)
rpart.plot(dem.cart)
```

---
### Random Forests

1.  Sample randomly from the entire data.
2.  Run CART, selecting a random subset of data points and variables at each decision node.
3.  Repeat several times, and find a way to average the results
    together.

---
```{r, echo = TRUE, out.width="90%", fig.retina = 1, fig.align='center'}
library(randomForest)
```

---
```{r, echo = TRUE, out.width="90%", fig.retina = 1, fig.align='center'}
dem.rf <- randomForest(democracy ~ corruption+genderequity+percapitagdp+refugees+centralbankindependence+imports+ecologicalfootprint+globalization, data=qogsmall, na.action=na.omit)

dem.rf
```

---
```{r, echo = TRUE, out.width="90%", fig.retina = 1, fig.align='center'}
library(randomForestExplainer)
```

---
```{r, echo = TRUE, out.width="90%", fig.retina = 1, fig.align='center'}
dem.mindepth <- min_depth_distribution(dem.rf)
plot_min_depth_distribution(dem.mindepth)
```


---
```{r, echo = FALSE, out.width="70%", fig.retina = 1, fig.align='center'}
plot_min_depth_distribution(dem.mindepth)
```

---
### What This Plot Means

* Variables that appear closer to the root (smaller minimal depth) are more important.

* This plot shows that corruption and per capita GDP are the strongest predictors of democracy in this model.

---

```{r, echo = TRUE, out.width="90%", fig.retina = 1, fig.align='center'}
plot_min_depth_interactions(dem.rf)
```

---

```{r, echo = FALSE, out.width="70%", fig.retina = 1, fig.align='center'}
plot_min_depth_interactions(dem.rf)
```

---
### What This Plot Means

This shows which pairs of variables frequently work together in the trees. For example, corruption and GDP might jointly predict democracy.

---
### What Is Machine Learning For?

* Especially prediction and forecasting: which kinds of countries are most likely to experience a coup next year?
* Can be a powerful tool for complex description problems: what topics dominate congressional speeches during conflicts?
* Researchers use machine learning to solve many kinds of challenges in causal inference: LASSO (a variant of regression) for variable selection, forests for estimating heterogeneous treatment effects
