---
title: "Midterm Review Session"
subtitle: "Political Science 210 — Fall 2026"
author: "Jaye Seawright"
date: "October 27, 2026 (in class)"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
---
class: center, middle

```{css, echo=FALSE}
pre {
  max-height: 400px;
  overflow-y: auto;
}
pre[class] {
  max-height: 200px;
}
```

```{r, setup, include=FALSE}
library(knitr)
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_mono_accent(
  base_color = "#1c5253",
  header_font_google = google_font("Josefin Sans"),
  text_font_google   = google_font("Montserrat", "300", "300i"),
  code_font_google   = google_font("Fira Mono"),
  text_font_size = "1.6rem"
)
```

### Today: Midterm Review

- **Format:** In‑class, open notes, closed devices. 75 minutes.
- **What to bring:** Your notes (handwritten or printed). No laptops/phones.
- **Coverage:** All material from Sept 24 through Oct 22 (up to multivariate regression).

We'll review **concepts and tools** using studies we've discussed in class.

---

## Example Studies We'll Use

| Study | Topic | Key concepts |
|-------|-------|--------------|
| Lott & Mustard (1996) | Concealed carry laws & crime | Four questions, confounding, reverse causation |
| Mask mandate study (Health Affairs) | COVID‑19 policies & infection rates | Controlling for confounders, difference‑in‑differences |
---

## Example Studies We'll Use

| Study | Topic | Key concepts |
|-------|-------|--------------|
| ANES regression example | Transgender bathrooms & carbon regulation | Bivariate/multivariate regression interpretation |
| **Piazza (2023)** | Political polarization & political violence | Observational research, cross‑national time‑series, correlation vs. causation, survey experiment |

---

## New Example: Piazza (2023)

**Research question:** Does affective partisan polarization increase support for political violence and the actual occurrence of political violence?

**Design:** Two studies:
1. **Original US survey** (N=1,899) measuring polarization (aversion to opposing party) and support for political violence.
2. **Cross‑national time‑series** of 83 democracies (1970–2018) testing whether polarized democracies experience more political violence.

---

**Findings:**
- In the US, polarized Democrats are 8% more likely, and polarized Republicans 18% more likely, to endorse political violence.
- In the cross‑national analysis, highly polarized democracies are 34% more likely to experience frequent political violence.

---

### Strengths and Weaknesses

| Aspect | Implication |
|--------|----------------|
| **Observational design** | No random assignment – confounding is a major threat. |
| **Survey** (US) | Measures polarization as an attitude, not a manipulated treatment. |
| **Cross‑national time‑series** | Uses lagged variables and controls (GDP, inequality, civil war history) to reduce confounding. |

---

### Strengths and Weaknesses

| Aspect | Implication |
|--------|----------------|| **Correlation vs. causation** | Piazza is careful: he finds an *association* and acknowledges possible reverse causation or common causes. |
| **External validity** | Findings from the US may not generalize to other democracies; cross‑national analysis helps but has its own limits. |

---

### Practice: Apply the four questions to Piazza

**Claim:** Polarization causes political violence.

- **Q1:** Is there a relationship? (Yes – both the US survey and cross‑national analysis show a positive association.)
- **Q2:** Could the outcome cause the treatment? (Possibly – political violence might increase polarization (rally effect).)
- **Q3:** Is there a causal pathway? (Piazza proposes dehumanization, moral polarization, and group mobilization.)

---

### Practice: Apply the four questions to Piazza

- **Q4:** Have confounders been ruled out? (Partially – he controls for many variables (GDP, inequality, past violence), but unmeasured confounders (e.g., elite rhetoric, social media) remain.)

**Takeaway:** Even with careful controls, observational studies cannot fully establish causation.

---

## The Four Questions (for causal inference)

| # | Question |
|---|----------|
| 1 | Is there a relationship between the treatment and the outcome? |
| 2 | Could the outcome cause the treatment? (reverse causation) |
| 3 | Is there evidence of a causal pathway? (mechanism) |
| 4 | Have confounding variables been ruled out? |

Use these to organize any critique of causal claims.

---

### Practice: Lott & Mustard

**What they claimed:** States that adopted “shall‑issue” concealed carry laws saw significant decreases in violent crime (murder ↓8.5%, rape ↓5%, aggravated assault ↓7%).

**Apply the four questions:**

- **Q1:** Is there a relationship?
- **Q2:** Could the outcome cause the treatment? 
- **Q3:** What is the proposed causal pathway? 
- **Q4:** Have confounders been ruled out?

---

### Practice: Confounding in Lott & Mustard

**Question:** Name **two potential confounders** that could bias the relationship between shall‑issue laws and crime rates. Explain why each meets the definition of a confounder.

*Think for a moment, then discuss with a neighbor.*

---

## Confounding and Control Variables

A **confounder** is a variable that causes **both** the treatment and the outcome.

**Controlling for a confounder** means holding it constant (e.g., including it in a regression) to isolate the causal effect of the treatment.

---

### Practice: Mask Mandate Study

From our causation lecture: a study comparing counties with and without mask mandates found that mandates were associated with lower COVID‑19 infection rates.

**Question:** What are some potential confounders? How might the authors try to control for them?

*Discuss.*

---

## Descriptive vs. Causal Inference

- **Descriptive:** What is the level or distribution of a variable? (e.g., “Shall‑issue states have higher homicide rates than may‑issue states.”)
- **Causal:** Does X cause Y? (e.g., “Shall‑issue laws cause higher homicide rates.”)

**Key:** Causal claims require a counterfactual and ruling out confounders.

---

### Practice: Descriptive or Causal?

| Claim | Type? |
|-------|-------|
| “States with shall‑issue laws have higher violent crime rates.” |   |
| “Shall‑issue laws increase violent crime.” |   |
| “Democrats are more likely to support mask mandates than Republicans.” |   |
| “Exposure to partisan news causes people to oppose mask mandates.” |   |
| “Polarized democracies are 34% more likely to experience political violence.” |   |
| “Polarization causes political violence.” |   |

---

## Regression Interpretation

### Bivariate Regression
$Y = \alpha + \beta X + \epsilon$

- $\beta$ = change in Y for a one‑unit change in X.
- p‑value for $\beta$ tests if the relationship is statistically significant.

---

## Regression Interpretation

### Multivariate Regression
Adds controls: $Y = \alpha + \beta_1 X_1 + \beta_2 X_2 + \dots + \epsilon$

- $\beta_1$ is the effect of $X_1$ **holding other variables constant**.
- If $\beta_1$ changes after adding a control, that control may be a confounder.

---

### Practice: ANES Regression Example

From our bivariate regression session: support for transgender bathroom laws (1–5 scale, higher = more conservative) predicts opposition to carbon regulations (1–5 scale, higher = more opposition).

**Bivariate result:** $\beta = 0.37$, p < 0.001.

---

### Practice: ANES Regression Example

**Multivariate result (adding party ID):** $\beta = 0.12$, p < 0.001; party ID coefficient = 0.21.

**Questions:**

1. Interpret the bivariate slope.
2. Interpret the multivariate slope for transbathroom.
3. What happened when party ID was added? What does that suggest?
4. Is party ID a confounder? Why?
---

## Observational Studies & Large‑N Methods

From the Piazza study: a cross‑national time‑series analysis of 83 democracies over nearly 50 years.

**Key features:**
- Observational (no randomization)
- Uses many countries and years (large N)
- Includes control variables (GDP, inequality, regime age, past violence, etc.)
- Finds that more polarized democracies are more likely to experience political violence.

---

## Observational Studies & Large‑N Methods

**Challenge:** Correlation is not enough to show causation. Even with many controls, unmeasured confounders (e.g., elite rhetoric, social media algorithms, historical legacies) may remain.

---

### Practice: Piazza Study – Alternative Explanations

**Question:** The Piazza study finds that polarized democracies have more political violence. Name **three alternative explanations** (other than polarization causing violence) and explain how you might test them.

*Think, then discuss.*

---

## Statistical Significance (p‑value)

- **p‑value:** Probability of observing the data (or more extreme) if the null hypothesis (no effect) were true.
- Small p‑value (< 0.05) → reject null → evidence of a relationship.
- Large p‑value → fail to reject null → not enough evidence.

**Important:** Statistical significance is not the same as practical importance.

---

### Practice: Interpret p‑values

From the Gadarian et al. study (discussed in class): Table 2 shows an adjusted risk difference for hand washing of 0.088 with p < 0.001.

- Is the difference statistically significant?
- What does the p‑value tell us? 
- Is the difference practically important? 

---

## Internal vs. External Validity

- **Internal validity:** Does the treatment cause the outcome for the cases studied?
- **External validity:** Do the results generalize to other populations, settings, or times?

**Trade‑off:** Often you have to sacrifice one for the other.

---

### Practice: Mask Mandate Study (again)

A study of U.S. counties in 2020 finds that mask mandates reduce infection rates.

- **Internal validity:** Threats? 
- **External validity:** Do the results apply to 2024? To other countries? 

---

## Sampling and Survey Methods

- **Random sample:** Every member of population has known, non‑zero chance – allows generalization.
- **Convenience sample:** Easy to access – may be biased.
- **Quota sample:** Match population characteristics on certain variables – better than convenience but not truly random.

---

### Practice: Sampling Bias

A study of vaccine acceptance uses an online survey recruited from Facebook ads. The sample is 10,000 people.

- Is this a random sample? 
- What biases might exist? 
- Could the results still be useful? 

---

## Measurement Validity

Does the measured variable actually capture the concept it claims to measure?

- **Example:** Measuring “political violence” using only terrorism deaths. Does it capture all political violence? 
- **Threats:** Social desirability bias, recall bias, ambiguous wording, missing dimensions.

---

### Practice: Measurement Validity in Piazza

**Question:** Piazza measures “polarization” using survey questions about whether the opposing party is a threat and whether one’s own party “loves America more.” Is this a valid measure of the concept of affective polarization? What might be missing?

*Discuss.*

---

## Final Tips for the Midterm

- **Read the instructions carefully.** Open notes, closed devices. Write legibly.
- **Use the four questions** to structure causal critiques.
- **When asked for a confounder,** explain why it meets the definition (causes both treatment and outcome).
- **When interpreting figures/tables,** look at confidence intervals and p‑values. Also explain what coefficients mean in substantive terms.
- **For measurement validity,** think about what the measure might miss or distort.
- **If you're stuck,** write something. Partial credit is better than blank.

---

### Good luck! Any final questions?

We'll now do a quick Q&A. Ask about concepts, not specific exam content.

After this, use the remaining time to review your notes quietly.
 studies. 