---
title: "Lab 1: Preparing for Homework 1"
subtitle: "Empirical Methods"
author: "<large>Jaye Seawright</large>"
institute: "<small>Northwestern Political Science</small>"
date: "Oct. 6, 2026"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
class: center, middle

```{css, echo=FALSE}
pre {
  max-height: 400px;
  overflow-y: auto;
}
pre[class] {
  max-height: 200px;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(ggplot2)
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_mono_accent(
  base_color = "#1c5253",
  header_font_google = google_font("Josefin Sans"),
  text_font_google   = google_font("Montserrat", "300", "300i"),
  code_font_google   = google_font("Fira Mono"),
  text_font_size = "1.6rem"
)
```

# Lab 1: Getting Ready for Homework 1

## Today's Goals

- Get comfortable with basic R operations
- Learn to work with two datasets at once
- Create variables and visualizations
- Connect these skills directly to Homework 1

---
class: inverse, center, middle

# Part 1: R Basics

---
### Getting Oriented

**R** does the calculations. **RStudio** is where we write and run code.

Four panes:
- **Source** (top left): where you write scripts
- **Console** (bottom left): where code runs
- **Environment** (top right): shows your data
- **Files/Plots/Help** (bottom right): where plots appear

---
### Loading Data with `read.csv()`

```{r, eval=TRUE}
# Read data from a URL
qog_small <- read.csv("https://github.com/jnseawright/ps210/raw/refs/heads/main/Data/qogsmall.csv")
```

**Key points:**
- `<-` is the assignment operator (saves data to a name)
- File paths matter! Use forward slashes `/` even on Windows

---
### Examining Data

```{r, eval=TRUE}
# Look at the first few rows
head(qog_small)
```

---
### Examining Data

```{r, eval=TRUE}
# Get summary statistics for all variables
summary(qog_small)
```

---
### Installing and Loading Packages

Packages extend R's capabilities.

```{r, eval=TRUE}
# Install a package (only need to do this once)
#install.packages("tidyverse")
#install.packages("ggplot2")

# Load a package (need to do this every session)
library(tidyverse)
library(ggplot2)
```

**Remember:** Install once, load every time!

---
class: inverse, center, middle

# Part 2: Working with Two Datasets

---
### Today's Example Data

We'll use data about democracy from the QOG dataset and right‑wing violence incidents:

```{r, eval=TRUE}
# Right-wing violence data (each row is one incident)
rtv <- read.csv("https://raw.githubusercontent.com/jnseawright/ps210/main/Data/rtv.csv")
```

---
### Examine Both Datasets

```{r, eval=TRUE}
head(rtv)
```

```{r, eval=TRUE}
head(qog_small)
```

**Questions:**
- What variables does `rtv` have?
- What variables does `qog_small` have?
- How are they connected? (What common variable could we use to match them?)

---
### Merging Datasets

We want to count how many violent incidents happened in each country‑year and add that count to the QOG data.

```{r, eval=TRUE}
# Step 1: Count incidents per country‑year
event_counts <- rtv %>%
  group_by(country_name, year) %>%
  summarise(rightwingviolence = n(), .groups = "drop")

# Step 2: Join to QOG
qog_merged <- left_join(qog_small, event_counts,
                        by = c("cname" = "country_name", "year" = "year"))

# Step 3: Replace NA with 0 (no incidents)
qog_merged$rightwingviolence <- ifelse(is.na(qog_merged$rightwingviolence), 0,
                                        qog_merged$rightwingviolence)

# Step 4: Keep only countries that appear in RTV (optional)
qog_merged <- qog_merged %>% filter(cname %in% rtv$country_name)
```

---
### Check Your Work

```{r, eval=TRUE}
head(qog_merged)
```

---
### Creating New Variables

We can create new variables based on existing ones:

```{r, eval=TRUE}
# Create a categorical variable from a numeric one
qog_merged$violence_cat <- cut(qog_merged$rightwingviolence,
                           breaks = c(-1, 0, 5, 10, Inf),
                           labels = c("None", "Little", "Some", "Much"))
table(qog_merged$violence_cat)
```

This is exactly the same logic as creating population groups in Homework 1.

---
class: inverse, center, middle

# Part 3: Visualization

---
### Scatter Plot of Democracy Level by Violence

```{r, eval=TRUE}
ggplot(qog_merged, aes(x = rightwingviolence, y = democracy)) +
  geom_point(alpha = 0.3) +
  labs(title = "Level of Democracy by Right-Wing Violence",
       x = "Right-Wing Violence",
       y = "Level of Democracy") +
  theme_minimal()
```

---
### What Does This Plot Tell Us?

- What is the relationship between right‑wing violence and democracy?
- What are the *limitations* of this plot?

---
### Grouped Bar Plot

Suppose we want to compare democracy scores between countries with "None" vs. "Little" violence, broken down by region.

```{r, eval=TRUE}
qog_merged$region <- sample(c("Europe", "Americas", "Asia", "Africa"),
                            size = nrow(qog_merged), replace = TRUE)
# (In real use, you would have a real region variable)

ggplot(qog_merged, aes(x = violence_cat, fill = region)) +
  geom_bar(position = "dodge") +
  labs(title = "Count of Country-Years by Violence Category and Region",
       x = "Violence Category",
       y = "Count",
       fill = "Region") +
  theme_minimal()
```

This is similar to the bar plot you will create in Homework 1, where you compare arrests by population group and governor's party.

---
### Key `ggplot` Concepts for Homework 1

- `aes(x = , fill = )` – map variables to x‑axis and fill color
- `geom_bar(position = "dodge")` – bars side‑by‑side (not stacked)
- `scale_fill_manual(values = c("blue", "red"))` – set custom colors
- `labs()` – add titles and axis labels

---
class: inverse, center, middle

# Part 4: Your Turn (20 min)

*Now open the Lab1_Practice.Rmd file and follow the instructions. We'll circulate to help.*

---
class: inverse, center, middle

# Part 5: Connecting to Homework 1

---
### How Today's Lab Maps to Homework 1

| Lab Skill | Homework 1 Problem |
|-----------|-------------------|
| Loading data with `read.csv()` | Problem 1: Load governor and ICE data |
| Examining data with `head()`, `summary()` | Problem 1: Understand what's in each dataset |
| Merging with `left_join()` | Problem 1: Add party variable to ICE data |
| Creating variables with `cut()` | Problem 3: Create population groups |
| Grouped bar plot with `ggplot()` | Problem 3: Visualize arrests by party and population |
| Interpreting limitations | Problems 2 and 4: Causal inference, experiment feasibility |

---
### Where to Find Help in Homework 1

- **Problem 1:** The code to merge is provided – you just need to run it and interpret
- **Problem 2:** Use the four questions from lecture (relationship, reverse causation, pathway, confounders)
- **Problem 3:** The population group code is provided – focus on interpreting the plot
- **Problem 4:** Think about what an experiment would require (random assignment, manipulation)

---
### Before You Leave

**Check that you can:**
- [ ] Load a CSV file into R
- [ ] Examine data with `head()` and `summary()`
- [ ] Merge two datasets using `left_join()`
- [ ] Create a new categorical variable with `cut()`
- [ ] Make a grouped bar plot with `ggplot()`
- [ ] Explain what the plot shows and its limitations

---
### Questions?

- TAs will be available in lab and office hours
- Get in touch by email or reach out to fellow students
- Don't wait until the night before!
