---
title: "Lab 1: Preparing for Homework 1"
subtitle: "Empirical Methods"
author: "<large>Jaye Seawright</large>"
institute: "<small>Northwestern Political Science</small>"
date: "Oct. 21, 2025"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false

---
class: center, middle

```{css, echo=FALSE}
pre {
  max-height: 400px;
  overflow-y: auto;
}

pre[class] {
  max-height: 200px;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(ggplot2)
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_mono_accent(
  base_color = "#1c5253",
  header_font_google = google_font("Josefin Sans"),
  text_font_google   = google_font("Montserrat", "300", "300i"),
  code_font_google   = google_font("Fira Mono"),
  text_font_size = "1.6rem"
)
```

# Lab 1: Getting Ready for Homework 1

## Today's Goals

- Get comfortable with basic R operations
- Learn to work with two datasets at once
- Create variables and visualizations
- Connect these skills directly to Homework 1

---
class: inverse, center, middle

# Part 1: R Basics

---
### Getting Oriented

**R** does the calculations. **RStudio** is where we write and run code.

Four panes:
- **Source** (top left): where you write scripts
- **Console** (bottom left): where code runs
- **Environment** (top right): shows your data
- **Files/Plots/Help** (bottom right): where plots appear

---
### Loading Data with `read.csv()`

```{r, eval=TRUE}
# Read data from a URL
qog_small <- read.csv("https://github.com/jnseawright/ps210/raw/refs/heads/main/Data/qogsmall.csv")

# Read data from your computer (after downloading)
# local_data <- read.csv("path/to/your/file.csv")
```

**Key points:**
- `<-` is the assignment operator (saves data to a name)
- File paths matter! Use forward slashes `/` even on Windows

---
### Examining Data

```{r, eval=TRUE}
# Look at the first few rows
head(qog_small)
```

---
### Examining Data

```{r, eval=TRUE}
# Look at the last few rows
tail(qog_small)
```

---
### Examining Data

```{r, eval=TRUE}
# Get summary statistics for all variables
summary(qog_small)
```

---
### Examining Data

```{r, eval=TRUE}
# See the structure (variable types)
str(qog_small)

# Open in spreadsheet view (capital V!)
#View(qog_small)
```

---
### Installing and Loading Packages

Packages extend R's capabilities.

```{r, eval=TRUE}
# Install a package (only need to do this once)
#install.packages("tidyverse")
#install.packages("ggplot2")

# Load a package (need to do this every session)
library(tidyverse)
library(ggplot2)
```

**Remember:** Install once, load every time!

---
### Quick Practice

With a neighbor, discuss:
1. What's the difference between `head()` and `summary()`?
2. Why do we need to load packages even after installing them?
3. What would happen if you run `summary(example_data)` before loading the data?

---
class: inverse, center, middle

# Part 2: Working with Two Datasets

---
### Today's Example Data

We'll use data about democracy from the QOG data that we loaded above, and also data about :

```{r, eval=TRUE}
# Right-wing violence data (each row is one incident)
rtv <- read.csv("https://raw.githubusercontent.com/jnseawright/ps210/main/Data/rtv.csv")
```

---
### Look at Both Datasets

```{r, eval=TRUE}
head(rtv)
```

---
### Look at Both Datasets
```{r, eval=TRUE}
head(qog_small)
```

---

**Questions:**
- What variables does `rtv` have?
- What variables does `qog_small` have?
- How are they connected? (What common variable could we use to match them?)

---
### Merging Datasets: The Loop Approach

Homework 1 uses a loop. Let's understand a slightly modified version:

```{r, eval=TRUE}
# Create variable with zero in it as defaul
qog_to_merge <- qog_small

qog_to_merge$rightwingviolence <- 0

# Loop through each act of violence
for (i in 1:nrow(rtv)) {
  # Find which country this act of violence happened in
  this_country <- rtv$country_name[i]
  this_year <- rtv$year[i]
  
  # Find the row(s) in qog_small that match this event's country and year
  target_rows <- qog_to_merge$cname == this_country & qog_to_merge$year == this_year

  # Increase the count for those rows by 1
  qog_to_merge$rightwingviolence[target_rows] <- qog_to_merge$rightwingviolence[target_rows] + 1
}

  #Our dataset only covers some countries.
qog_plus_rtv <- qog_to_merge %>% filter(cname %in% rtv$country_name)
qog_to_merge <- NULL
```

---
### A Faster and Shorter Alternative

The `tidyverse` gives us a speedier and less error-prone way:

```{r, eval=TRUE}
# Step 1: Count events per country-year from the events dataset
event_counts <- rtv %>%
  group_by(country_name, year) %>%
  summarise(rightwingviolence = n(), .groups = "drop")

# Step 2: Join these counts to the full QoG panel
qog_plus_rtv <- left_join(qog_small, event_counts, 
                          by = c("cname" = "country_name", "year" = "year"))

# Step 3: Replace NA counts with 0 (countries/years with no events)
qog_plus_rtv$rightwingviolence <- ifelse(is.na(qog_plus_rtv$rightwingviolence), 0, 
                                          qog_plus_rtv$rightwingviolence)

# Step 4: Filter countries not in both datasets.
qog_plus_rtv <- qog_plus_rtv %>% filter(cname %in% rtv$country_name)
```

---
### A Faster and Shorter Alternative

This does the same thing as the loop, but it will execute much more quickly!

---
### Check Your Work

```{r, eval=TRUE}
# See what we have now
head(qog_plus_rtv)
```

---
### Check Your Work

```{r, eval=TRUE}
# How much violence?
# Load ggplot2
library(ggplot2)

# Create a histogram of violence per country-year
rwhist <- qog_plus_rtv %>% ggplot(aes(x = rightwingviolence)) +
  geom_histogram(bins = 10,           # number of bins (adjust as needed)
                 fill = "skyblue",    # bar fill color
                 color = "white") +   # bar outline color
  labs(title = "Distribution of Right-Wing Violence",
       x = "Violence Per Country-Year",
       y = "Count") +
  theme_minimal()                     # clean theme
```

---

```{r, eval=TRUE}
rwhist
```

---
### Creating New Variables

We can create new variables based on existing ones:

```{r, eval=TRUE}
# Create a variable indicating large states
qog_plus_rtv$loggedrightwingviolence <- log(qog_plus_rtv$rightwingviolence + 1)

rwloghist <- qog_plus_rtv %>% ggplot(aes(x = loggedrightwingviolence)) +
  geom_histogram(bins = 10,           # number of bins (adjust as needed)
                 fill = "skyblue",    # bar fill color
                 color = "white") +   # bar outline color
  labs(title = "Distribution of Logged Right-Wing Violence",
       x = "Logged Violence Per Country-Year",
       y = "Count") +
  theme_minimal()                     # clean theme
```

---

```{r, eval=TRUE}
rwloghist
```

---

```{r, eval=TRUE}
# Create a categorical variable from a numeric one
qog_plus_rtv$violence_cat <- cut(qog_plus_rtv$rightwingviolence,
                           breaks = c(-1, 0, 5, 10, Inf),
                           labels = c("None", "Little", "Some", "Much"))
table(qog_plus_rtv$violence_cat)
```

---
### Your Turn (with a neighbor)

Given what we've seen:
1. Why might we prefer the tidyverse approach over writing a loop?
2. What could go wrong if the country (or state?) names don't match exactly?
3. How would you check if the merge worked correctly?

---
class: inverse, center, middle

# Part 3: Visualization

---
### Scatter Plot of Democracy Level by Violence

```{r, eval=TRUE}
rwscatter <- qog_plus_rtv %>% ggplot(aes(x = rightwingviolence, y = democracy)) +
  geom_point() +
  labs(title = "Level of Democracy by Right-Wing Violence",
       x = "Right-Wing Violence",
       y = "Level of Democracy") +
  theme_minimal()
```


---

```{r, eval=TRUE}
rwscatter
```

---
### What Does This Plot Tell Us?

Look at the plot and discuss:

- What is the relationship between right-wing violence and democracy?
- What are the *limitations* of this plot?

---
### Visualizing Logged Violence Instead

Maybe we don't like the stretched distribution of violence, so let's use the log:

```{r, eval=TRUE}
rwlogscatter <- qog_plus_rtv %>% ggplot(aes(x = log(1+ rightwingviolence), y = democracy)) +
  geom_point() +
  labs(title = "Logged Level of Democracy by Right-Wing Violence",
       x = "Logged Acts of Right-Wing Violence",
       y = "Level of Democracy") +
  theme_minimal()
```

---

```{r, eval=TRUE}
rwlogscatter
```

---
### Does this help?

Look at the plot and discuss:

- Does the log transformation make the relationship easier or harder to see?
- What are the *advantages* and *disadvantages* of this version of the plot?

---
class: inverse, center, middle

# Part 4: Your Turn (20 min)

*Now you'll practice with a different dataset. Open the Lab1_Practice.Rmd file and follow the instructions. We'll circulate to help.*

---
class: inverse, center, middle

# Part 5: Connecting to Homework 1 

---
### How Today's Lab Maps to Homework 1

| Lab Skill | Homework 1 Problem |
|-----------|-------------------|
| Loading data with `read.csv()` | Problem 1: Load governor and ICE data |
| Examining data with `head()`, `summary()` | Problem 1: Understand what's in each dataset |
| Combining two datasets | Problem 1: Add party variable to ICE data |

---
### How Today's Lab Maps to Homework 1

| Lab Skill | Homework 1 Problem |
|-----------|-------------------|
| Creating variables with `mutate()` and `cut()` | Problem 3: Create population groups |
| Bar plots with `ggplot()` | Problem 3: Visualize arrests by party and population |
| Interpreting limitations | Problems 2 and 4: Causal inference, experiment feasibility |

---
### Key Differences to Notice

| Today's Lab | Homework 1 |
|-------------|------------|
| Democracy and violence | ICE arrests data |
| tidyverse (faster) | Loop (more explicit) |
| Party from `states` dataset | Party from `governor_data` |
| Population groups with quintiles | Same approach! |

The *logic* is the same even if the code differs slightly.

---
### Where to Find Help in Homework 1

- **Problem 1:** The loop code is already provided—you just need to run it and interpret the results
- **Problem 2:** Use the four questions from lecture (relationship, reverse causation, pathway, confounders)
- **Problem 3:** The code for population groups is provided—focus on interpreting the plot
- **Problem 4:** Think about what an experiment would require (random assignment, manipulation)

---
### Before You Leave

**Check that you can:**
- [ ] Load a CSV file into R
- [ ] Examine data with `head()` and `summary()`
- [ ] Merge two datasets using `left_join()`
- [ ] Create a new categorical variable with `cut()`
- [ ] Make a bar plot with `ggplot()`
- [ ] Explain what the plot shows and its limitations

---
### Questions?

- TAs will be available in lab and office hours
- Get in touch by email or reach out to fellow students
- Don't wait until the night before!